---
title: "LiRA - WG1 Vector non EU from EFSA Systematic literature review"
author: "Alfredo Acosta SVA"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
    toc_float: true
    theme: cosmo
    fig_caption: yes
    number_sections: yes
    code_folding: show
  word_document: default
  html_notebook: 
    toc: yes
---
# Analisis of efsa publication on vectors
- Systematic literature review on the vector status of potential 
vector species of 36 vector-borne pathogens
- Information needed for the integration of WG1-WG2

```{r }
# Library ----
library(readxl)
library(dplyr)
library(lubridate)
library(leaflet)
library(sf)
library(DT)
library(ggplot2)
```

```{r}
setwd("C:/Users/alfredo.acosta/OneDrive - SVA/LORA/Working example/vectors/")
```

# DB EfSA
## Reading data
```{r warning=FALSE}
w <- read_xlsx("DL3.1.2.SLR_VectorStatus_Database_comp.xlsx", sheet = 1)
```
# Contents of the DB
```{r}
colnames(w)
table(w$year)
table(w$Continent)
table(w$Host)
table(w$VBD_AS_Full)
table(w$VBD_Agent_subtype)
table(w$VBD_Agent_subtype)
table(w$VectorGroup)
table(w$VectorSpecies)
length(unique(w$VectorSpecies))
head(w)
```

# Species specified on the Vector net output
```{r}
species <- c("Aedes albopictus",
             "Aedes vexans vexans",
             "Aedes aegypti",
             "Aedes japonicus",
             "Culex pipiens",
             "Culex modestus", #MC: 28.11.2023 added
             "Culex tarsalis", ## not in the dataset
             "Culex quinquefasciatus") ## not in the dataset

table(w$VectorSpecies %in% species)

```
# Number of reports for every country
```{r}
reports <- w %>% 
  group_by(Country, Country_ISO) %>% 
  summarise(n(), diseases=length(unique(VBD_AS_Full)), vectorSP=length(unique(VectorSpecies)))

```
## Reading the World map
```{r}
wm <- st_read("TM_WORLD_BORDERS_SIMPL-0.3.shp")

```

## Number of reports on vectors (pop-up showing the number of on the map
```{r}
  # WORLD MAP ----
  # Update the number of outbreaks on the map
  wm$reports <- reports$`n()`[match(wm$ISO2, reports$Country_ISO)]
  wm$diseases <- reports$diseases[match(wm$ISO2, reports$Country_ISO)]
  wm$vectors <- reports$vectorSP[match(wm$ISO2, reports$Country_ISO)]

  str(wm$reports)

# Maps of outbreaks
  pal <- colorNumeric(
    palette = "Spectral",
    domain = wm$reports, 
    na.color = "white"
  )


# Maps of cases ----
pal <- colorNumeric(
  palette = "Spectral",
   reverse = TRUE,
  domain = wm$reports, 
  na.color = "white")

map <- leaflet(wm) %>%
  addProviderTiles(providers$CartoDB.Positron)

mc <- map %>%
  setView(lng = 25, lat = 10, zoom = 2) %>%
  addPolygons(weight = 0.5, 
              color = ~pal(reports),
              popup = ~paste("Diseases:", diseases, "<br>", "Vectors", vectors)) %>% 
  addLegend("bottomright", pal = pal, values = ~reports,
            # title = paste("Cases",
            #               "<br>",
            #               date_to_analyse, 
            #               "<br>", 
            #               disease_to_analyse),
            opacity = 1)

mc

```

# African Swine fever
```{r}
reports <- w %>% 
  filter(VBD_Agent_subtype == "ASFV") %>% 
  group_by(Country, Country_ISO) %>% 
  summarise(n(), vector=length(unique(VectorSpecies)), vectorSP=paste(unique(VectorSpecies), collapse = ", "))
reports
```

## Map Number of reports on vectors (pop-up showing the name of on the map)
```{r}
  # WORLD MAP ----
  # Update the number of outbreaks on the map
  wm$reports <- reports$`n()`[match(wm$ISO2, reports$Country_ISO)]
  wm$vector <- reports$vector[match(wm$ISO2, reports$Country_ISO)]
  wm$vectorSP <- reports$vectorSP[match(wm$ISO2, reports$Country_ISO)]

  str(wm$reports)

# Maps of outbreaks
  pal <- colorNumeric(
    palette = "Spectral",
    domain = wm$reports, 
    na.color = "white"
  )


# Maps of cases ----
pal <- colorNumeric(
  palette = "Spectral",
   reverse = TRUE,
  domain = wm$reports, 
  na.color = "white")

map <- leaflet(wm) %>%
  addProviderTiles(providers$CartoDB.Positron)

mc <- map %>%
  setView(lng = 25, lat = 10, zoom = 2) %>%
  addPolygons(weight = 0.5, 
              color = ~pal(reports),
              popup = ~paste("N vectors:", vector, "<br>", "Names", vectorSP)) %>% 
  addLegend("bottomright", pal = pal, values = ~reports,
            # title = paste("Cases",
            #               "<br>",
            #               date_to_analyse, 
            #               "<br>", 
            #               disease_to_analyse),
            opacity = 1)

mc

```


# West nile virus
```{r}
reports <- w %>% 
  filter(VBD_Agent_subtype == "WNV") %>% 
  group_by(Country, Country_ISO) %>% 
  summarise(n(), vector=length(unique(VectorSpecies)), vectorSP=paste(unique(VectorSpecies), collapse = ", "))
reports
```

## Map Number of reports on vectors (pop-up showing the name of on the map)
```{r}
  # WORLD MAP ----
  # Update the number of outbreaks on the map
  wm$reports <- reports$`n()`[match(wm$ISO2, reports$Country_ISO)]
  wm$vector <- reports$vector[match(wm$ISO2, reports$Country_ISO)]
  wm$vectorSP <- reports$vectorSP[match(wm$ISO2, reports$Country_ISO)]

  str(wm$reports)

# Maps of outbreaks
  pal <- colorNumeric(
    palette = "Spectral",
    domain = wm$reports, 
    na.color = "white"
  )


# Maps of cases ----
pal <- colorNumeric(
  palette = "Spectral",
   reverse = TRUE,
  domain = wm$reports, 
  na.color = "white")

map <- leaflet(wm) %>%
  addProviderTiles(providers$CartoDB.Positron)

mc <- map %>%
  setView(lng = 25, lat = 10, zoom = 2) %>%
  addPolygons(weight = 0.5, 
              color = ~pal(reports),
              popup = ~paste("N vectors:", vector, "<br>", "Names", vectorSP)) %>% 
  addLegend("bottomright", pal = pal, values = ~reports,
            # title = paste("Cases",
            #               "<br>",
            #               date_to_analyse, 
            #               "<br>", 
            #               disease_to_analyse),
            opacity = 1)

mc

```

# BTV
```{r}
reports <- w %>% 
  filter(VBD_Agent_subtype == "BTV") %>% 
  group_by(Country, Country_ISO) %>% 
  summarise(n(), vector=length(unique(VectorSpecies)), vectorSP=paste(unique(VectorSpecies), collapse = ", "))
reports
```

## Map Number of reports on vectors (pop-up showing the name of on the map)
```{r}
  # WORLD MAP ----
  # Update the number of outbreaks on the map
  wm$reports <- reports$`n()`[match(wm$ISO2, reports$Country_ISO)]
  wm$vector <- reports$vector[match(wm$ISO2, reports$Country_ISO)]
  wm$vectorSP <- reports$vectorSP[match(wm$ISO2, reports$Country_ISO)]

  str(wm$reports)

# Maps of outbreaks
  pal <- colorNumeric(
    palette = "Spectral",
    domain = wm$reports, 
    na.color = "white"
  )


# Maps of cases ----
pal <- colorNumeric(
  palette = "Spectral",
   reverse = TRUE,
  domain = wm$reports, 
  na.color = "white")

map <- leaflet(wm) %>%
  addProviderTiles(providers$CartoDB.Positron)

mc <- map %>%
  setView(lng = 25, lat = 10, zoom = 2) %>%
  addPolygons(weight = 0.5, 
              color = ~pal(reports),
              popup = ~paste("N vectors:", vector, "<br>", "Names", vectorSP)) %>% 
  addLegend("bottomright", pal = pal, values = ~reports,
            # title = paste("Cases",
            #               "<br>",
            #               date_to_analyse, 
            #               "<br>", 
            #               disease_to_analyse),
            opacity = 1)

mc

```

# MIR Mininum infection rates 
## Selection of pririotized diseases
```{r}
# ASF	African Swine Fever
# CCHF	Crimean-Congo Haemorrhagic fever
# EHD	Epizootic Haemorrhagic Disease
# EE	Equine Encephalosis
# *** HPAI	Highly Pathogenic Avian Influenza
# JE	Japanese Encephalitis 
# *** LSD	Lumpy Skin Disease
# *** Rab	Rabies
# RVF	Rift Valley Fever
# WNF	West Nile Fever
# 
head(w)

# Which abreviattion correspond to the disease
w %>% 
group_by(w$VBD_Agent_subtype, w$VBD_AS_Full) %>% 
  summarise(n())

di <- c("ASFV","BTV","CCHFV", "EHDV", "EEV", "JEV", "RVFV", "WNV")

```

## Options to filter
```{r}
table(w$fedStatus)
table(w$LifeStage)
```
## Filtering criteria for Mosquitoes, Bitting midges and Sand flies (127)
```{r}
a <- w %>% 
  filter(MIR > 0) %>% 
  # filter(VBD_Agent_subtype %in% di) %>%
  filter(fedStatus == "Unfed" | fedStatus == "Gravid") %>% 
  filter(VectorGroup == "Mosquitoes" | VectorGroup == "Biting midges" | VectorGroup == "Sand flies") %>%
  filter(LifeStage == "Female")
datatable(a)
```

## Filter criteria for Ticks (27)
```{r}
b <-  w %>% 
    filter(MIR > 0) %>% 
    filter(fedStatus == "Unfed" | fedStatus == "Gravid") %>% 
    filter(VectorGroup == "Ticks")

  datatable(b)
```
## Number of MIR values (Checking filter)
```{r}
# Minimum infection rates
# After applying the filtering criteria on the feeding status and life stage, 42 MIR values were
# found or calculated for mosquitoes, 67 for sand flies, 27 for ticks, and 18 for biting midges.
# Records including MIR values are visualized in a boxplot grouped by vector group (Figure 4).
# Details per arthropod species are provided in Appendix F.

c <- rbind(a,b)
c %>% group_by(VectorGroup) %>% 
  summarise(n())

```

# MIR Table
```{r}
c <- rbind(a,b)

datatable(c %>% 
  filter(VBD_Agent_subtype %in% di) %>%
  group_by(VBD_AS_Full, VectorGroup) %>% 
  summarise(obs = n(),Min=round(min(MIR)), Median=round(median(MIR),2), Max=round(max(MIR),2)))

```


## MIR considerig disease, vector group and species
```{r}
datatable(c %>% 
  filter(VBD_Agent_subtype %in% di) %>%
  group_by(VBD_AS_Full, VectorGroup, VectorSpecies) %>% 
  summarise(obs = n(),Min=round(min(MIR)), Median=round(median(MIR),2), Max=round(max(MIR),2)))


```

## Boxplot of MIR - Disease
```{r}
c %>% 
  filter(VBD_Agent_subtype %in% di) %>%
  group_by(VBD_AS_Full, VectorGroup) %>% 
  ggplot(aes(MIR, VBD_AS_Full)) +
  geom_boxplot()+
  labs(y = "Disease", x= "MIR %o")
```

## Boxplot of MIR - Species
```{r}
c %>% 
  filter(VBD_Agent_subtype %in% di) %>%
  group_by(VBD_AS_Full, VectorGroup, VectorSpecies) %>% 
  ggplot(aes(MIR, color=VBD_AS_Full, VectorSpecies)) +
  geom_boxplot()+
  facet_grid(cols=vars(VBD_AS_Full), scales = "free")+
  theme(legend.position="none")+
  labs(y = "Disease", x= "MIR %o")


```

# Credits 
Acosta, Alfredo PhD<sup>1</sup>. 

**Swedish Veterinary Agency SVA<sup>1</sup>**: SVA <http://www.sva.se/>. 